---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
# Reproducible Research: Peer Assessment 1
## Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks.  

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Loading and preprocessing the data

```{r, echo=TRUE, options(scipen = 4, digits = 4)}
datafile <- unzip("activity.zip")
data <- read.csv(datafile)
```
The variables included in this dataset are:  
__steps__: Number of steps taking in a 5-minute interval (missing values are coded as NA)  
__date__: The date on which the measurement was taken in YYYY-MM-DD format  
__interval__: Integer identifier for the 5-minute interval in which measurement was taken

The dataset has a total of 17,568 observations in three variables.
```{r, echo=TRUE}
str(data)
```

## What is mean total number of steps taken per day?
To get a first impression of the sort of data we are working on we start with the average number of steps the sample person takes every day (disregarding NA values) and we draw a histogram.  
1. Calculate the total number of steps taken per day.  
```{r, echo=TRUE}
steps.sum <- tapply(data$steps, data$date, sum)
print(steps.sum)
```
2. Make a histogram of the total number of steps taken each day:
```{r, echo=TRUE, fig.height=4}
hist(steps.sum, main="Number of steps per day", xlab="Number of steps")
```

3. Calculate and report the mean and median of the total number of steps taken per day.  
```{r, echo=TRUE}
summary(steps.sum)
mean.steps <- mean(steps.sum, na.rm=T)
median.step <- median(steps.sum, na.rm=T)
```
The mean is `r mean.steps` and the median is `r median.step`.

## What is the average daily activity pattern?
Now we look at the average daily activity, that is how many steps we take during the day, in the morning, for example, or at night?  
4. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis):  
```{r, echo=TRUE, fig.height=4}
steps.interval <- tapply(data$steps, data$interval, mean, na.rm=T)
intervals <- strptime(sprintf("%04d", as.numeric(names(steps.interval))), format="%H%M")
plot(intervals, steps.interval, type="l", main="Average number of steps during the day", xlab="Interval", ylab="Average number of steps")
```
  
5. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo=TRUE}
steps.interval[order(steps.interval, decreasing = TRUE)][1]
```
where the first value is the time of the day and the second the number of steps.


## Imputing missing values  
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.  
6. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs):
```{r, echo=TRUE}
s <- sum(is.na(data$steps))
```
The total number of missing values is `r s`.  
7. Devise a strategy for filling in all of the missing values in the dataset.  
The strategy will be putting in place of NA the average number of steps across all intervals in the dataset.  
```{r econ}
# generating a new copy
data2 <- data
mean.na <- mean(steps.sum, na.rm=TRUE)/288
data2$steps <- ifelse(is.na(data2$steps), mean.na, data2$steps)
```

The value that will substitute every NA is `r mean.na`.  
8. Create a new dataset that is equal to the original dataset but with the missing data filled in:  
The new dataset will be:
```{r, echo=TRUE}
head(data2)
```
9. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
This is the histogram
```{r, echo=TRUE, fig.height=4}
steps2.sum <- tapply(data2$steps, data2$date, sum)
hist(steps2.sum, main="Total number of steps taken each day", xlab="Daily steps")
smr <- summary(steps2.sum)
mean2.steps <- mean(steps2.sum)
median2.step <- median(steps2.sum)
```
The new data are the following and they not differ very much from the previous histogram in which NA were present: mean = `r mean2.steps`; median = `r median2.step`.  
The impact of imputing missing data is eliminating zero values in the night hours. It should be important to analyse further the question.  






## Are there differences in activity patterns between weekdays and weekends?

9. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r, echo=TRUE, warning=FALSE}
library("dplyr")
data2$date <- as.Date(data2$date)
data.final <- select(data2, steps, date, interval)
data.final <- mutate(data.final, weekday = weekdays(date))

dayVect <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
data.final <- data.final %>% mutate(wdays = ifelse(data2$weekday %in% dayVect, "weekday", "weekend"))
data.final$wdays <- factor(c("weekday", "weekend"))
```
10. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).  
```{r, echo=TRUE}

library("lattice")
data.final <- group_by(data.final, interval, wdays)
mean.steps.wkd <- summarize(data.final, msteps=mean(steps))
xyplot(msteps ~ interval | wdays, data = mean.steps.wkd, type="l", main="Average number of steps\nacross all weekday days or weekend days", ylab="Average Number of Steps", xlab="Interval", layout = c(1,2))
```
